\LARGE \bf Approximately Orchestrated Road Traffic Automata\\ a General Traffic Simulation Framework

### 2012-03-07 papers/itsc_2012/draft2.tex
paper edits, filling out sections in simulation especially

\section{ARCHITECTURE}

@@

AORTA is divided into three modular components: the map model, a
micro-simulation engine, and a UI. The map model transforms OSM maps into AORTA
graphs, then answers pathfinding and geometry queries. The simulation engine
adds a notion of agents, vehicle dynamics, and collisions. A headless mode
enables running experiments without the overhead of visualization. Finally, the
UI, which uses Swing, can interactively render the map and agents. For
convenience, specific traffic patterns can be constructed between two regions of
the map by simply drawing a polygon around each set of roads.

AORTA's implementation uses Scala, a language implemented on the Java Virtual
Machine. Scala provides the advantage of functional programming constructs while
still permitting imperative style. Parts of AORTA, including extensions and clients, can
just as easily be written in Java. The software has a documented API, is
open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
the code for the experiments described in Section 6 -- it is designed for
extensibility.

@@

\section{MAP CONSTRUCTION FROM OSM}

Our goal is to simulate traffic on existing infrastructure, not contrived,
randomly generated maps. We do this by utilizing OpenStreetMaps, a community
project offering user-editable, free maps. To run a simulation on a map from
OSM, there is a one-time conversion process to transform the map into AORTA's
format, which is also encoded with XML.

@@

A map is represented as a directed graph with \emph{lanes} as edges and
\emph{turns} as vertices. A \emph{road} groups one or two groups of lanes
together, each group being a different direction of travel.
\emph{Intersections} map the possible turns from incoming to outgoing lanes.
Since both lanes and turns support traffic, they are grouped together as the
\emph{traversables}, meaning they have some geometric total length.

The map also has a geometric interpretation, used for visualization and physical
distances. A road has a sequence of points giving the line segments of the
center yellow line. These points could also be interpreted as control points for
a curved road, but AORTA uses straight lines as an approximation. Lanes are
parallel to the center line, some width apart from one another.  Turns are
modeled with a single straight line connecting the end of one edge to the start
of another (so right turns often have $0$ length).

@@

\commentd{gotta work on passive voice}

Map construction is split into several passes, each feeding the next.
\begin{enumerate}
  \item OSM encodes many paths besides driveable roads, so these are first
        filtered out. GPS locations common to multiple roads are marked, since
        they implicitly indicate intersections.
  \item OSM \emph{ways} (sequences of GPS locations) include many
        intersections, so the builder next splits ways into undirected
        segments of roads between exactly two intersections.
  \item Undirected roads are multiplied into directed lanes in each direction
        (unless the road is marked one-way). The number of lanes is guessed from
        OSM's ``road type'' tag or an explicit number of lanes, when given.
  \item Turns are next constructed between incoming and outgoing edges at each
        vertex. Left, right, straight, and U-turns are heuristically chosen
        based on the angles between edges. When several lanes all cross into
        fewer lanes, merging is arbitrarily forced on the rightmost lanes.
        Although inaccurate, these patterns are the best compromise. At the
        time of writing, even sources like Google Maps (TODO cite) lack this
        information.
  \item Because turns are heuristically placed, the graph is disconnected. The
        strongly connected components of the digraph with the fewest members are
        removed. Tarjan's algorithm locates these in linear time (TODO cite).
        Disconnected lanes are spliced out of larger roads, leaving unappealing
        geometric gaps, but correctly ensuring connectivity for corner-case free
        pathfinding.
  \item The builder serializes the graph in a simple XML format.
\end{enumerate}

@@

\section{SIMULATION}

AORTA uses microscopic simulation, modeling individual drivers as agents.
Discrete-time simulation is used, meaning an agent accelerates at a fixed rate
for the duration of a ``tick'' (also referred to as ``step'') to achieve a new
position and velocity. The tick duration $dt$ is fixed, enabling agents to
reason about collision avoidance and satisfying other constraints. Setting this
value to the average ``reaction time'' of a human driver may be appropriate for
some experiments. Each tick, the simulation performs the following steps:

\commentd{space continuous? get terminology right}

\begin{enumerate}
  \item Introducing new agents into the map
  \item Updating the position and velocity of agents based on their choice the
        previous tick
  \item Checking for collisions
  \item Allowing each agent to react
\end{enumerate}

Agents have a few enforced constraints -- they cannot exceed their physical
acceleration capabilities, travel in reverse, or travel past the end of a
traversable without appearing on the next. Intersection policies and avoiding
collisions on the same traversable provide the other constraints, but it is the
job of an agent's behavior to satisfy those. There are three main components of
the simulation, each configurable by extending default implementations of an
interface: agent behaviors, routing strategies, and intersection policies.

%\commentd{justify arbitrary constants, like deaccel for cars;
%          systems ideas, or limitations.. trylocks, atomic sim update, deadlock
%          issues;
%          limit: no lane-changing}

\subsection{Spawning}

To introduce new agents into the simulation in real-time, users create
\emph{generators} programatically or using the UI. A generator picks a random
starting position and random goal from two sets of roads chosen by the user,
selectable by drawing a polygon with the mouse in the UI. A generator may cover
the entire map, uniformly distributing traffic everywhere, or it can mimic
rush-hour scenarios.

Every tick, a generator creates some number of new agents. If the agent's route
policy needs work to be done (such as pathfinding), the generator either
performs it immediately (incurring a delay in simulation if it is running) or
sends it to a pool of worker threads to compute in the background.

Once an agent's route is ready, the agent waits alongside its starting edge (as
if it is in a driveway or parking lot). When the edge has no agents that could
potentially crash with the new vehicle, the new agent enters the system with
zero initial speed.  Because behaviors must enforce speed limit along an edge by
the time an agent enters the road, then if there is no agent within the
worst-case distance (traveling distance at the speed limit, plus stopping
distance from this speed) before the spawn point, then the new agent may enter.
To render unnecessary the prediction of which agents could enter the edge in the
near future, there is a buffer space at the beginning of the edge where a new
agent may never spawn. This disqualifies some short edges from being spawn
points, but this is not a big limitation in practice.

@@

All traversables (edges and the straight line-approximation of turns) maintain
an ordered queue of occupying agents. Agents stepping forward in time happens
atomically, so collision checking occurs after all agents have moved for some
tick. This is done because otherwise false positives could occur when an agent
steps before the agent in front of it, and ordering agents for updates is
difficult near the crossings between edges. \commentd{Do I need to explain why
it is difficult, or should I just stop after ``in front of it''?} A collision is
defined as the reversal of order of agents before and after a step.
\commentd{lazy queue checking for speedup. mention briefly?}

@@

Each agent has a \emph{behavior} governing it. When the agent travels past the end
of an edge during its step, the behavior picks the turn to pursue. Each tick,
the behavior can make the agent react by performing one of two actions:
disappearing from the map (only when the agent is at rest and is done with its
route) or setting an acceleration for the next step. Were lane changing or
merging modeled, that would be another possible action.

Currently, the primary behavior used by all agents is the ``route-following
behavior.'' It is a generalized, baseline behavior that guarantees not to
collide with another agent or enter an intersection at the wrong time. Since it
picks the highest safe choice of acceleration at each step, it could be easily
extended to mimic human drivers by traveling at some random speed slower than
this or to optimize fuel efficiency by tuning acceleration.

\subsubsection{Worst-case analysis and lookahead}

@@

The default behavior maximizes acceleration while obeying speed limits,
avoiding collisions, and obeying intersections. By considering the highest speed
achievable at the end of the next tick and the subsequent stopping distance at
this speed, the behavior can scan sufficiently far ahead along several
traversables to locate agents to avoid and intersections to query. The analysis
can short-circuit at the first nearby agent and intersection requiring a stop,
since it would be illegal to pass either.

\subsubsection{Interaction with intersections}

% TODO regroup? does this belong as a subsection, not subsub?

A behavior's interaction with an intersection policy amounts to polling it with
the turn the agent wants to perform, the distance away the agent currently is,
and how long the agent has been waiting (so policies can enforce a required
delay before entering the intersection). The intersection policy simply orders
the agent to ``stop'' or ``continue.'' The acceleration to avoid entering an intersection
attempts to end a configurable threshold back from the end of a traversable to
account for crosswalks.

Intersection policies are one of the most intriguing aspects of the simulation
to tweak, especially in light of autonomous vehicles. Current policies include
traditional stop signs and traffic signals. Future policies could include stop
lights that adaptively learn better timings and an AIM-based reservation system
\cite{JAIR08-dresner}.

\commentd{Actually, at least the stop sign policy is really mundane. Include its
  description?}

%The stop sign policy works by maintaining a current owner and a queue of waiting
%agents. An agent can enter the queue only when it is within some threshold of
%the start of its turn. After the agent at the head of the queue has waited for a
%configurable pause, it obtains the lock and has exclusive access to the
%intersection, releasing it only when it completes its turn.

\commentd{describe traffic signal policy, once it's working fully. the delay
thing is maybe interesting.}

@@

Picking accelerations that optimize some objective functions is a separate task
from picking a path to travel. Behaviors consult a route strategy to pick turns.
Currently two simple implementations are available. One statically plans a route
to some goal using standard A* search (TODO cite); the other performs a drunken
walk, lazily planning random choices at each intersection as the lookahead
engine needs to know. Future experiments in dynamic replanning or hierarchial
planning can be performed without modifying any code other than a creating new
route implementation. Route policies could also interact with a singleton
overseer to have some sort of global insight.

\subsection{Deadlock}

Maintaining following distance from agents actually enables a form of deadlock
in the simulation. If an agent is forced to stop in an intersection mid-turn due
to its destination edge filling up with other agents, then nobody may cross the
blocked intersection. With many agents in a small area, this situation may
occur several times in interdependent ways. The solution currently being
investigated will require behaviors to ensure a turn may be fully completed
before starting it. This will prevent intersections from ever being blocked.

\commentd{this is now fixed in most cases. describe, and describe a better
  workaround for some cases: S turns}
\commentd{and describe the full-capacity case and how that's hopeless anyway}

%\commentp{Cite (and ideally use ideas from) paper commented out in latex file
    % @inproceedings{AAAI11-au,
    % Author = {Tsz-Chiu Au and Neda Shahidi and Peter Stone},
    % Booktitle= {Proceedings of the Twenty-Fifth Conference on Artificial Intelligence},
    % Title = {Enforcing Liveness in Autonomous Traffic Management},
    % Year = {2011},
    % Month= {August},
%}

% TODO solve the above case, then figure out what to describe.

%A form of deadlock is possible in a sequence of short edges and turns. Agent
%$A_1$ acquires the lock to intersection $V_1$, but needs the lock to $V_2$ to
%proceed. Meanwhile, $A_2$ has the lock for $V_2$ and needs the lock for $V_1$.
%To deal with this circumstance, an agent's behavior can yield any held locks (a
%notion generalized to any intersection policy) when its speed is $0$ and its
%conservative acceleration is to remain at rest. The stop sign policy only
%removes an agent's lock if it has not started its turn (because otherwise, the
%agent is stalling in the intersection). The next time that agent polls, the
%policy restores the agent to the front of the queue.

%\commentd{hopefully we'll have a neater solution than the above soon}

@@

AORTA is designed for controlled experiments. Consequently, simulations are
deterministically reproducible. A single seed for the pseudo-random number
generator, the input graph, and the generators' configuration completely
determine the outcome of any simulation. This determinism does break down when
generators use worker threads, since routes may be computed in different orders,
causing agents to be ready to enter the road at different times.

@@

\section{EXPERIMENTAL RESULTS}

@@

\section{CONCLUSIONS}

@@

\section*{ACKNOWLEDGMENT}

>>>>

\section{ARCHITECTURE}
\label{sec:arch}

@@

AORTA is divided into three modular components: the map model, micro-simulation
engine, and UI. The map model transforms OSM maps into AORTA graphs, then
answers pathfinding and geometry queries. The simulation engine adds a notion of
agents, vehicle dynamics, and collisions. Finally, the UI, which uses Swing, can
interactively render the map and agents. A headless mode also enables running
experiments without the overhead of visualization.

AORTA's implementation uses Scala, a language implemented on the Java Virtual
Machine. Scala provides the advantage of functional programming constructs while
still permitting imperative style. Parts of AORTA, including extensions and
clients, can easily be written in Java or Scala. The software, designed for
extensibility, has a documented API, is
open-source\footnote{http://code.google.com/p/road-rage}, and contains all of
the code for the experiments described in Section \ref{sec:results}.

@@

\section{MAP CONSTRUCTION FROM OSM}
\label{sec:map}

Our goal is to simulate traffic on existing infrastructure, not contrived,
randomly generated maps. Utilizing OpenStreetMaps, a community project offering
user-editable, free XML-based maps, enables this. To run a simulation on a map
from OSM, there is a one-time conversion process to transform the map into
AORTA's format, which is also encoded with XML.

@@

A map is represented as a directed graph with \emph{lanes} as edges and
\emph{turns} as vertices. A \emph{road} groups one or two groups of lanes
together, each group being a different direction of travel.
\emph{Intersections} map the possible turns from incoming to outgoing lanes.
Since both lanes and turns support traffic, they are grouped together as
\emph{traversables}, meaning agents may exist on them.

The map also has a geometric interpretation, used for visualization and physical
distances. A road contains a sequence of points giving the line segments of the
center yellow line. These points could also be interpreted as control points for
a curved road, but AORTA uses straight lines as an approximation. Lanes are
parallel to the center line, some fixed width apart from one another. Turns are
modeled with a single straight line connecting the end of one lane to the start
of another (so right turns often have $0$ length).

@@

Map construction is split into several passes, each feeding the next.
\begin{enumerate}
  \item OSM encodes many paths besides driveable roads, so these are first
        filtered out. GPS locations common to multiple roads are marked, since
        they implicitly indicate intersections.
  \item OSM \emph{ways} (sequences of GPS locations) include many
        intersections, so the builder next splits ways into undirected
        segments of roads between exactly two intersections.
  \item Undirected roads are multiplied into directed lanes in each direction
        (unless the road is marked one-way). The number of lanes is guessed from
        OSM's ``road type'' tag or an explicit number of lanes, when given.
  \item Turns are next constructed between incoming and outgoing lanes at each
        vertex. Left, right, straight, and U-turns are heuristically chosen based on
        the angles between lanes. When several lanes all cross into fewer lanes,
        merging is arbitrarily forced on the rightmost lanes.  Although inaccurate,
        these patterns are the best compromise. At the time of writing, even sources
        like Google Maps \commentd{TODO cite} lack this information.
  \item Because turns are heuristically placed, the graph is usually
        disconnected. Since the largest resulting map is desirable, the strongly
        connected components of the digraph with the fewest members are removed.
        Disconnected lanes are spliced out of larger roads, leaving unappealing
        geometric gaps, but correctly ensuring connectivity for corner-case free
        pathfinding.
  \item The builder serializes the graph in a simple XML format.
\end{enumerate}

@@

\section{SIMULATION}
\label{sec:simulation}

AORTA uses microscopic simulation, modeling individual drivers as agents.
Simulation is discrete-time, meaning an agent accelerates at a fixed rate for
the duration of a ``step'' to achieve a new position and velocity, and
space-continuous, meaning roads are not tiled. The step duration $dt$ is fixed,
enabling agents to reason about collision avoidance and satisfying other
constraints. Setting this value to the average ``reaction time'' of a human
driver may be appropriate for some experiments. Each step, the simulation
performs the following:

\begin{enumerate}
  \item Introducing new agents into the map
  \item Updating the position and velocity of agents based on their choice the
        previous step
  \item Checking for collisions
  \item Allowing each agent to react
\end{enumerate}

Agents have a few enforced constraints -- they cannot exceed their physical
acceleration capabilities, travel in reverse, or travel past the end of a
traversable without appearing on the next. Intersection policies and avoiding
collisions on the same traversable provide the other constraints, but the
agent's behavior must satisfy these. There are three main components of the
simulation, each configurable by extending built-in implementations of an
interface: agent behaviors, routing strategies, and intersection policies.

Currently lane-changing is not supported. Merging lanes is accomplished by
treating the merge point as an intersection.

\subsection{Spawning new agents}

To introduce new agents into the simulation in real-time, users can create
\emph{generators} programatically or using the UI. A generator picks a random
starting position and goal from two sets of roads. These roads can be selected
by mouse with a bounding polygon in the UI, letting specific traffic patterns
like rush-hour scenarios easily be constructed between two regions. A generator
may also cover the entire map, uniformly distributing traffic everywhere.

Every step, a generator creates some number of new agents. If the agent's route
policy needs computation like pathfinding to operate, the generator either
performs it immediately (incurring a delay in simulation if it is running) or
sends it to a pool of worker threads to process in the background.

Once an agent's route is ready, the agent waits alongside its starting lane (as
if it is in a driveway or parking lot). When the lane has no agents that could
potentially crash with the new vehicle, the new agent enters the system with
zero initial speed.  Because behaviors must enforce speed limit along a lane 
from the moment an agent enters the road, then if there is no agent within the
worst-case distance (traveling distance at the speed limit, plus stopping
distance from this speed) before the spawn point, then the new agent may enter.
To avoid predicting which agents might enter the lane in the near future, there
is a buffer space at the beginning of the lane where a new agent may never
spawn. This disqualifies some short lanes from being spawn points, but this is
not a big limitation in practice.

@@

All traversables (lanes and the straight line-approximation of turns) maintain
an ordered queue of occupying agents. One simulation step happens atomically,
eaning collision checking occurs after all agents have moved. If this was not
done, false positives could occur when an agent steps before the agent in front
of it. A collision is defined as the reversal of order of agents before and
after a step.

@@

Each agent has a \emph{behavior} governing it. When the agent travels past the end
of a lane during its step, the behavior picks the turn to pursue. Each step,
the behavior can make the agent react by performing one of two actions:
disappearing from the map (only when the agent is at rest and is done with its
route) or setting an acceleration for the next step. Were lane changing or
merging modeled, that would be another possible action.

Currently, the primary behavior used by all agents is a generalized, baseline
behavior that guarantees not to collide with another agent or enter an
intersection at the wrong time. Since it picks the highest safe choice of
acceleration at each step, it could be easily extended to mimic human drivers by
traveling at some random slower acceleration or to optimize fuel efficiency
by tuning movement.

@@

The default behavior maximizes acceleration while obeying speed limits,
avoiding collisions, and obeying intersections. By considering the highest speed
achievable at the end of the next step and the subsequent stopping distance at
this speed, the behavior can scan sufficiently far ahead along several
traversables to locate agents to avoid and intersections to query. The analysis
short-circuits at the first nearby agent and intersection requiring a stop,
since it would be illegal to pass another anyway.

\subsection{Intersection Policies}

A behavior's interaction with an intersection policy amounts to polling it with
the turn the agent wants to perform, the distance away the agent currently is,
and how long the agent has been waiting (so policies can enforce a required
delay before entering the intersection). The intersection policy simply orders
the agent to ``stop'' or ``proceed.'' The acceleration chosen to avoid entering
an intersection attempts to account for crosswalks by ending a configurable
threshold before the end of a traversable.

Intersection policies are one of the most intriguing aspects of the simulation
to tweak, especially in light of autonomous vehicles. Current policies include
traditional stop signs, traffic signals, and an AIM-inspired autonomous
reservation system \cite{JAIR08-dresner}. The traffic signals are further
extensible by assigning groups of non-conflicting turns at each intersection
with some duration and offset. The experiment in section \ref{sec:results}
features standard turn grouping (turns from parallel incoming directions are
scheduled simultaneously, and left turns are grouped) and a breadth-first search
heuristic for propagating timings. Future policies could include signals that
adapt timings and refinements to the autonomous reservation policy.

@@

Picking accelerations that optimize objective functions is a separate task from
picking a path to travel. Agent behaviors consult a route strategy to pick
turns. Currently two simple implementations are available. One statically plans
a route to some goal using standard A* search (\commentd{TODO cite}); the other
performs a drunken walk, lazily planning random choices at each intersection as
the lookahead engine needs to know. Future experiments in dynamic replanning or
hierarchial planning can be performed without modifying any code other than a
creating new route implementation. Route policies could also interact with a
singleton overseer to gain some sort of global insight or with smart
intersections to avoid congestion.

\subsection{Gridlock}

Agents maintain a safe following distance from the next agent. If a lane is
filled, then an agent may be forced to stop in an intersection mid-turn and
block other traffic. The default behavior prevents this by refusing to start a
turn before the lookahead engine guarantees no agent will trigger a premature
stop.

\commentd{wiki has a page naming gridlock. cite that for the name?}

However, with very high volumes of traffic, it is possible for waiting agents to
accumulate and fill lanes to their capacity. When this happens in a circular
manner so that the agent at the front of each lane is waiting to perform a turn
into another full lane with the same such head agent, gridlock occurs; no agent
in the system will make progress unless an agent at the front decides to pick a
different turn. At best, AORTA can detect this by maintaining a dependency graph
that encodes the ``is following'' relation and finding cycles. It is not clear
how to prevent or mend the deadlock without forcing agents to re-route.

\commentd{is the above clear and an acceptable limitation? should i find other
sources that have the same troubles?}

@@

AORTA is designed for controlled experiments. Consequently, simulations are
deterministically reproducible. A seed for the pseudo-random number generator,
the input graph, and the generators' configuration determine the outcome of any
simulation. This determinism may break down when generators use worker threads,
since routes may be computed in different orders, causing agents to enter the
road at different times.

@@

\section{EXPERIMENTAL RESULTS}
\label{sec:results}

@@

\section{CONCLUSIONS}
\label{sec:conclusion}

@@

\section*{ACKNOWLEDGMENTS}
